How to select predictive models for decision-making or causal inference?

### 2024-08-29 response_to_reviewer.md

Another problem is that the paper is hard to follow. The authors rapidly switch between topics: risk, selection, nuisance parameter estimation so the reader quickly gets lost.  The sections in supplementary materials contain important information but not all are referenced in main text.  The paper gives an impression of several experiments put together in some order without guiding the reader through the maze.  I think organization should be improved before publication.

Other comments:
---------------

In the Methods section g-computation is described.  It seems indeed the first name under which the technique appeared in literature, but other names, most importantly S-learner are being used frequently nowadays.  Interestingly those names are mentioned in Empirical study on page 4, however, the close relationship (equivalence?) between S-learner and g-computation is not mentioned.

> G-computation is larger than S-learner. The first mention of S-learner that we are aware of is [Künzel, 2019](https://arxiv.org/pdf/1706.03461). In this paper, they distinguish T-learners from S-learners (or more elaborated ones). T-learners model $\mu(0)$ and $\mu(1)$ separately and train separately on each population. On the contrary S learner use only one model for both population and add the treatment as a covariate. A good example to understand the difference is to note that a linear T-learner is the same thing as a linear S-learner where we add an interaction term between the treatment and the covariates. Ultimately, both methods are G-computation in the sense that they rely on the identification of eq. (1) making the link between potential outcomes and observed quantities. To improve the manuscript, we added just after the eq. (3) that one can choose to model the response function with either a T-learner or a S-learner.

@@

> Thanks. Fixed.

A.7: the data generation procedure needs better explanations.  Why not call (7) P(d|A) since that's how its used later?  On page 17 you say '[b1,bD]~P(x)' and later 'x~P(x)' which causes confusion.  Also, how are beta vectors generated?

page 4: TLearner uses different feature representation for each treatment arm.  Does this really happen in practice?  Usually all subgroups are described with the same set of features.

> It is indeed true that all subgroups are described with the same set of features. However,

@@

page 4/5: Family of candidate estimators: why are only linear models used in some dataset and forests for other?  The leaf sizes of forests are chosen from a rather small range, why not use some percentage of the size of the training set?

> The choice of the candidate family of estimators was guided by the idea to compare estimators that are not too different from each other, making the model selection task more challenging. The important point is that we are not focusing on the best possible estimator but on the best possible model selection procedure. To identify the optimal procedure, it is more convincing to compare close candidate estimators than a wide variety of candidate estimators. In the later case, one estimator could dominate all other estimators for every selection procedure. This explains why we are choosing a narrow range of leaf sizes for the forests.

> For the simulations we chose estimators that are out of the functional class of the  true data generation process but are quite similar to it. They are out of the funtional class because the representers are different from the one of the true data. They are close because they have the same functional form : a linear model on top of representers. We used the simulation to explore the wider range of data generation processes, focusing on a gradient of overlap. Because of the sheer number of generated dataset, we did not include computationnaly costly boosted trees in the candidate family.  than the one used for the data generation process. An important point is that the . For the semi-simulated datasets, we chose . For a forest estimator, a T learner or a S learner are quite similar . The fact that the .

### 2024-08-29 response_to_reviewer.md

page 4: TLearner uses different feature representation for each treatment arm.  Does this really happen in practice?  Usually all subgroups are described with the same set of features.

> It is indeed true that all subgroups are described with the same set of features. However,

page 4/5: Family of candidate estimators: why are only linear models used in some dataset and forests for other?  The leaf sizes of forests are chosen from a rather small range, why not use some percentage of the size of the training set?

> For the simulations we chose estimators that are out of the functional class of the  true data generation process (a candidate estimator cannot model exactly the data generation process) but that are quite similar to it. They are out of the functional class because the representers are different from the one of the true data (they are two randomly sampled data points). They are close because they have the same functional form : a linear model on top of two data representers. We used the simulations to explore the wider range of data generation processes, focusing on a gradient of overlap. Because of the sheer number of generated dataset, we did not include computationally costly boosted trees in the candidate family.

### 2024-08-29 response_to_reviewer.md

I am however not sure about the overall conclusions, which are that R-risk is always the best.  Performance of risk functions may depend on several factors and the overall best performance might be a result of bias in available data.  The only factor taken into account by the authors is overlap.  However, only one type of lack of overlap is tested in experiments, based on shifted Gaussians.  This will likely lead to violation of the positivity assumption.  How about different scenarios, e.g.: perfect overlap but different treatment/control distributions?  Confounding need not necessarily lead to lack of overlap.  Another possible factor is the ratio of causal effect to background responses.  It is known that it may affect relative performance of HTE estimators.

Overall: I am no fully convinced that R-risk is always the best.  The answer could be dependent on other factors and I hope the authors would explore more scenarios.

Another question: do the differences between risks in experiments follow known theoretical results?  For example, it is well known (e.g. https://www.semanticscholar.org/reader/112aba9c62f98b9684693f4c20c30274bc9eb1dc or [Rudas ans jaroszewisz, 2017](https://link.springer.com/content/pdf/10.1007/s10618-018-0576-8.pdf)) that the tau risk performs poorly.

> The differences between risks have been explored *theoretically* by XX for the lose performance of the U-risk. For the other risk, we are not aware of any theoretical results. 
> The differences between risks have been highlighted *experimentally* in other papers such as [Schuler, et al., 2018](https://arxiv.org/pdf/1804.05146) for quadratic and stepwise data generation processes or [Curt et Van der Schaar, 2023](https://proceedings.mlr.press/v202/curth23b/curth23b.pdf) exploring more diverse outcome generation processes such as CATE complexity. They both underline good performances for the R risk.

@@

> Thanks for this remark. We fixed it.

A.7: the data generation procedure needs better explanations.  Why not call (7) P(d|A) since that's how its used later?  On page 17 you say '[b1,bD]~P(x)' and later 'x~P(x)' which causes confusion.  Also, how are beta vectors generated?

page 4: TLearner uses different feature representation for each treatment arm.  Does this really happen in practice?  Usually all subgroups are described with the same set of features.

> It is indeed true that in usual practice all subgroups are described with the same set of features. However, as soon as we are using a machine learning (necessary to model CATE) such as a random forest or boosted trees, the inner representations of the features inside the model is different for both population. This idea has been exploited to build representation aware neural net models for cate estimation by [Johansson et al., 2022](https://www.jmlr.org/papers/volume23/19-511/19-511.pdf). It is also the case for simpler models such as a linear model with treatment interaction and lasso regularization.

@@

Algorithm 1: 'a candidate estimators f' should be 'a candidate estimator f'

throughout: "an RBF expansion" is more commonly used than "a RBF expansion"

page 4: 'for both population' should be 'for both populations'

> Thanks for the remarks. We corrected the typos and improved the language as suggested.

### 2024-08-29 how_to_select_latex.tex
wip: review iterate

\begin{frontmatter}
    \maketitle
    \begin{abstract}
        \textbf{Background:}
        We investigate which procedure selects the predictive model most
        trustworthy to reason on the effect of an intervention and support
        decision making.\\
        %As predictive models --\eg from machine learning-- give likely
        %outcomes, they may be used to, a
        %causal-inference task. The increasing complexity of health data has opened
        %the door to a plethora of models, but also the Pandora box of model
        %selection: which of these models yield the most valid causal estimates? 
        \textbf{Methods:}
        We study such a variety of model selection procedures in practical settings:
        finite samples settings and without theoretical assumption of
        \emph{well-specified} models. Beyond standard cross-validation or
        internal validation procedures, we also study elaborate causal risks.
        These build proxies of the causal error using ``nuisance''
        re-weighting to compute it on the observed data. We evaluate whether
        empirically estimated nuisances, which are necessarily noisy,
        add noise to model selection. We compare different metrics for
        causal model selection in an extensive empirical
        study based on a simulation and three healthcare datasets
        based on real covariates. \\
        \textbf{Results:} Among all metrics, the mean squared error,
        classically used to evaluate predictive modes,
        is worse. Re-weighting it with propensity score
        does not bring much improvements. The
        $R\text{-risk}$, which uses as nuisances a model of
        mean outcome and propensity scores, leads to the best performances.
        Nuisance corrections are best estimated with flexible estimators such
        as a super learner.
        \\
        \textbf{Conclusions:} When predictive models are used to
        reason on the effect of an intervention, they must be evaluated
        with different procedures than standard predictive
        settings; using the $R\text{-risk}$ from causal inference.
    \end{abstract}

@@

Indeed, concluding on the effect of a treatment is a difficult
causal-inference task, as it can be easily compromised by confounding:
spurious associations between treatment
allocation and baseline health, \eg~only prescribing a drug to mild cases
\cite{hernan_causal_2020,vanderweele2019principles}.
Predictive modeling bridges to causal inference theory under the name of
\emph{outcome models} (or G-computation,
G-formula \cite{robins_role_1986}, Q-model
\cite{snowden_implementation_2011}, conditional mean regression
\cite{wendling_comparing_2018}).
% Controlled allocation of treatment, as in randomized controlled trials (RCTs),
% alleviate this concern. Yet most machine-learning models are trained on
% \emph{observational} data, close to real-word practice
% \cite{black1996we,hernan_methods_2021} but challenging for causal inference.
Medical statistics and epidemiology have mostly used other
causal-inference methods, modeling treatment assignment
with propensity scores \cite{rosenbaum_central_1983,austin_moving_2015,casucci2018estimating,grose_use_2020}. Outcome modeling brings
the benefit of going
beyond average effects, estimating individualized or conditional average
treatment effects (CATE), central to precision medicine.
%
For this purpose, such methods are also invaluable for randomized trials
\cite{su2018random,lamont2018identification,hoogland2021tutorial}.

@@

We first illustrate the problem of causal model selection. Then we anchor causal model selection in the \emph{potential outcome}
framework and details the causal risks and model-selection procedure. We then
rewrite the so-called $R\text{-risk}$ as a reweighted version of mean squared
difference between the true and estimated individualized treatment effect.
Finally, we conduct a thorough empirical study comparing the different metrics
on diverse datasets, using a family of simulations and real health
data, going beyond prior work limited to specific simulation settings
\cite{schuler_comparison_2018,alaa_validating_2019} (Appendix
\ref{apd:prior_work}).

@@

\begin{algorithm}[!htbp]
    \caption{Model selection procedure}\label{problem:estimation_procedure:algo} {%
        Given train and test sets $(\mathcal{T}, \mathcal{S}) \sim \mathcal{D}$,
        a candidate estimators $f$, a causal
        metrics $\ell$:
        \begin{enumerate}
            \item Prefit: Learn estimators for unknown nuisance quantities $(\check e,\,\check m)$ on the training set $\mathcal{T}$
            \item Fit: learn $\hat f(\cdot, a)$ on
                  $\mathcal T$
            \item Model selection: $\forall{x} \in \mathcal{S}$ predict
                  $\big(\hat f(x, 1), \hat f(x, 0)\big)$ and evaluate the estimator storing the metric value: $\ell(f, \mathcal S)$ -- possibly
                  function of $\check e$ and $\check m$
                  %\item Metric evaluation: return the oracle evaluation metrics
                  %evaluated on $\mathcal{S}$: $\big(\widehat{\tau\text{-risk}}(\hat
                  %f^*_{\ell}); \widehat{\ell}_{ATE}(\hat f^*_{\ell}) \big)$
        \end{enumerate}

@@

We test model selection across different candidate estimators that
approximate imperfectly the
data-generating process. To build such estimators, we first use a RBF
expansion similar to that used for data generation. We choose two random
knots and transform the raw data features with a Gaussian
kernel. This step is referred as the featurization. Then, we fit a linear
regression on these transformed features. We consider two ways of combining
these steps for outcome model; we use common nomenclature
\cite{kunzel_metalearners_2019,shen2023rctrep} to refer to these different
meta-learners that differ on how they model, jointly or not, the treated and the
non treated:
\begin{itemize}
    \item SLearner: A single learner for both population, taking the treatment as
          a supplementary covariate.
    \item SftLearner: A single set of basis functions is sampled at random for both
          populations, leading to a given feature space used to model both the treated and
          controls, then two
          separate different regressors are fitted on this shared representation.
    \item TLearner: Two completely different learners for each population, hence
          separate feature representations and regressors.
\end{itemize}

>>>>

\begin{frontmatter}
    \maketitle
    \begin{abstract}
        \textbf{Background:}
        We investigate which procedure selects the predictive model most
        trustworthy to reason on the effect of an intervention and support
        decision making.\\
        %As predictive models --\eg from machine learning-- give likely
        %outcomes, they may be used to, a
        %causal-inference task. The increasing complexity of health data has opened
        %the door to a plethora of models, but also the Pandora box of model
        %selection: which of these models yield the most valid causal estimates? 
        \textbf{Methods:}
        We study a large variety of model selection procedures in practical settings:
        finite samples settings and without theoretical assumption of
        \emph{well-specified} models. Beyond standard cross-validation or
        internal validation procedures, we also study elaborate causal risks.
        These build proxies of the causal error using ``nuisance''
        re-weighting to compute it on the observed data. We evaluate whether
        empirically estimated nuisances, which are necessarily noisy,
        add noise to model selection. We compare different metrics for
        causal model selection in an extensive empirical
        study based on a simulation and three healthcare datasets
        based on real covariates. \\
        \textbf{Results:} Among all metrics, the mean squared error,
        classically used to evaluate predictive modes,
        is worse. Re-weighting it with propensity score
        does not bring much improvements. The
        $R\text{-risk}$, which uses as nuisances a model of
        mean outcome and propensity scores, leads to the best performances.
        Nuisance corrections are best estimated with flexible estimators such
        as a super learner.
        \\
        \textbf{Conclusions:} When predictive models are used to
        reason on the effect of an intervention, they must be evaluated
        with different procedures than standard predictive
        settings; using the $R\text{-risk}$ from causal inference.
    \end{abstract}

@@

Indeed, concluding on the effect of a treatment is a difficult
causal-inference task, as it can be easily compromised by confounding:
spurious associations between treatment
allocation and baseline health, \eg~only prescribing a drug to mild cases
\cite{hernan_causal_2020,vanderweele2019principles}.
Predictive modeling is linked to causal inference theory by the concept of
\emph{outcome models} (or G-computation,
G-formula \cite{robins_role_1986}, Q-model
\cite{snowden_implementation_2011}, conditional mean regression
\cite{wendling_comparing_2018}).
% Controlled allocation of treatment, as in randomized controlled trials (RCTs),
% alleviate this concern. Yet most machine-learning models are trained on
% \emph{observational} data, close to real-word practice
% \cite{black1996we,hernan_methods_2021} but challenging for causal inference.
Medical statistics and epidemiology have mostly used other
causal-inference methods, modeling treatment assignment
with propensity scores \cite{rosenbaum_central_1983,austin_moving_2015,casucci2018estimating,grose_use_2020}. Outcome modeling brings
the benefit of going
beyond average effects, estimating individualized or conditional average
treatment effects (CATE), central to precision medicine.
%
For this purpose, such methods are also invaluable for randomized trials
\cite{su2018random,lamont2018identification,hoogland2021tutorial}.

@@

We first illustrate the problem of causal model selection. Then we anchor causal model selection in the \emph{potential outcome}
framework and detail the causal risks and model-selection procedure. We then
rewrite the so-called $R\text{-risk}$ as a reweighted version of mean squared
difference between the true and estimated individualized treatment effect.
Finally, we conduct a thorough empirical study comparing the different metrics
on diverse datasets, using a family of simulations and real health
data, going beyond prior work limited to specific simulation settings
\cite{schuler_comparison_2018,alaa_validating_2019} (Appendix
\ref{apd:prior_work}).

@@

\begin{algorithm}[!htbp]
    \caption{Model selection procedure}\label{problem:estimation_procedure:algo} {%
        Given train and test sets $(\mathcal{T}, \mathcal{S}) \sim \mathcal{D}$,
        a candidate estimator $f$, a causal
        metrics $\ell$:
        \begin{enumerate}
            \item Prefit: Learn estimators for unknown nuisance quantities $(\check e,\,\check m)$ on the training set $\mathcal{T}$
            \item Fit: learn $\hat f(\cdot, a)$ on
                  $\mathcal T$
            \item Model selection: $\forall{x} \in \mathcal{S}$ predict
                  $\big(\hat f(x, 1), \hat f(x, 0)\big)$ and evaluate the estimator storing the metric value: $\ell(f, \mathcal S)$ -- possibly
                  function of $\check e$ and $\check m$
                  %\item Metric evaluation: return the oracle evaluation metrics
                  %evaluated on $\mathcal{S}$: $\big(\widehat{\tau\text{-risk}}(\hat
                  %f^*_{\ell}); \widehat{\ell}_{ATE}(\hat f^*_{\ell}) \big)$
        \end{enumerate}

@@

We test model selection across different candidate estimators that
approximate imperfectly the
data-generating process. To build such estimators, we first use an RBF
expansion similar to that used for data generation. We choose two random
knots and transform the raw data features with a Gaussian
kernel. This step is referred as the featurization. Then, we fit a linear
regression on these transformed features. We consider two ways of combining
these steps for outcome model; we use common nomenclature
\cite{kunzel_metalearners_2019,shen2023rctrep} to refer to these different
meta-learners that differ on how they model, jointly or not, the treated and the
non treated:
\begin{itemize}
    \item SLearner: A single learner for both populations, taking the treatment as
          a supplementary covariate.
    \item SftLearner: A single set of basis functions is sampled at random for both
          populations, leading to a given feature space used to model both the treated and
          controls, then two
          separate different regressors are fitted on this shared representation.
    \item TLearner: Two completely different learners for each population, hence
          separate feature representations and regressors.
\end{itemize}
